---
title: "Lab 00"
#author: "Dr. Patrick Bitterman"
#date: "`r format(Sys.time(), '%Y %B %d')`"
output: pdf_document
my_subtitle: "Env Analysis in R"
fancy: true
geometry: margin=1in
latex_engine: pdflatex
colorlinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Lab 00: Getting set up with R, RStudio, and GitHub

### Read the instructions COMPLETELY before starting the lab

This lab is designed to get you set up with the tools we will use throughout the semester: R, RStudio, and GitHub. It also introduces you to the most basic operations in R. Completing this lab successfully means you have a working environment and can demonstrate some fundamental R skills. Everything in this lab will be used — and built upon — in every lab that follows.

### Formatting your submission

This lab must be placed into a public repository on GitHub (www.github.com). Before the due date, submit **on Canvas** a link to the repository. I will then download your repositories and run your code. For this lab, your submission must be an **RMarkdown (.Rmd) document**. Any data you reference must use **relative path names** so that your code runs on my machine as well as yours.


## Part 1: Getting set up

Work through each step below before moving on to the R content.

**Step 1: Install R**

Download and install R from CRAN: https://cran.r-project.org. Choose the version appropriate for your operating system. Accept all default settings during installation.

**Step 2: Install RStudio**

Download and install RStudio Desktop (the free version) from https://posit.co/download/rstudio-desktop. RStudio is an IDE (integrated development environment) that makes working with R much easier. R and RStudio are separate programs — you need both.

**Step 3: Create a GitHub account**

If you do not already have one, create a free account at https://github.com. Use a professional username, as you may use this account beyond this class.

**Step 4: Install GitHub Desktop**

Download and install GitHub Desktop from https://desktop.github.com. This application gives you a graphical interface for working with Git and GitHub without needing to use the command line.

**Step 5: Create a new repository**

On GitHub (in your browser), create a new **public** repository. Give it a sensible name (e.g., `env-analysis-labs`). Initialize it with a README file. This is the repository where all of your lab submissions will live this semester.

**Step 6: Clone the repository**

In GitHub Desktop, clone the repository you just created to a location on your computer that you can find easily. This downloads a local copy that stays synchronized with GitHub.

**Step 7: Open this lab file in RStudio**

Move or save this `.Rmd` file into your cloned repository folder. Open RStudio, then open the `.Rmd` file. You are now ready to begin.


## Part 2: Your first RMarkdown document

An RMarkdown document has three components:

1. **The YAML header** — the block at the top between the `---` lines. It controls metadata like the title and output format.

2. **Prose** — regular text written in Markdown. This is where you write explanations, answer questions, and describe your work.

3. **Code chunks** — blocks of R code enclosed between ` ```{r} ` and ` ``` `. R executes this code when you knit the document.

To produce a PDF from this document, click the **Knit** button at the top of the RStudio editor (or use Ctrl+Shift+K / Cmd+Shift+K). Try it now before continuing.

### Inserting a code chunk

You can insert a new code chunk by:

- Clicking the **Insert** button in the RStudio editor toolbar and selecting "R"
- Using the keyboard shortcut: Ctrl+Alt+I (Windows) or Cmd+Option+I (Mac)

Try inserting a chunk now and typing `1 + 1` inside it. Knit the document again and confirm the output appears in the PDF.


## Part 3: Basic R operations

R can be used as a calculator. The basic arithmetic operators are `+`, `-`, `*`, `/`, and `^` (exponent).

```{r arithmetic}
5 + 3
12 / 4
2^8
```

You can store values in **variables** using the assignment operator `<-`. Variable names are case-sensitive.

```{r variables}
x <- 42
y <- 3.14
x * y
```

R has several basic **data types**. The most common are:

- `numeric` — numbers (e.g., `3.14`, `100`)
- `character` — text, always in quotes (e.g., `"hello"`)
- `logical` — `TRUE` or `FALSE`

You can check the type of any object with `class()`:

```{r data-types}
class(42)
class("hello")
class(TRUE)
```


## Part 4: Vectors

A **vector** is an ordered collection of values of the same type. Create one with `c()` (short for "combine"):

```{r vectors}
temps <- c(72, 68, 75, 80, 65)
names <- c("Alice", "Bob", "Carol")
```

Arithmetic on numeric vectors is applied element-wise:

```{r vector-math}
temps + 10        # add 10 to every element
temps * (5/9)     # convert Fahrenheit to Celsius (roughly)
```

Several built-in functions summarize numeric vectors:

```{r vector-summary}
sum(temps)
mean(temps)
length(temps)
min(temps)
max(temps)
```


## Part 5: Packages

R's base functionality can be extended with **packages** — collections of functions written by the R community. Two important commands:

- `install.packages("packagename")` — downloads and installs the package. You only need to do this **once** per machine.
- `library(packagename)` — loads the package into your current R session. You need to do this **every time** you start a new session.

```{r load-packages, message=FALSE}
# install.packages("tidyverse")  # run this once in the console, not in your script
library(tidyverse)
```

Note that the `install.packages()` line above is commented out. It is generally bad practice to include `install.packages()` calls in a script or `.Rmd` file that will be run repeatedly or shared, because it re-installs the package every time the code runs.


## Part 6: Reading and summarizing data

We will start with R's built-in `iris` dataset, which requires no external file. You can load it with `data()`:

```{r load-iris}
data(iris)
```

Several functions help you get a quick look at any dataset:

```{r explore-iris}
head(iris)      # first 6 rows
tail(iris)      # last 6 rows
summary(iris)   # summary statistics for each column
glimpse(iris)   # compact overview of structure
```

In this course, you will more commonly read data from external files. The `read_csv()` function from the `tidyverse` is the standard way to do this. Note the use of a **relative path** — the `..` moves up one directory level from the `labs/` folder to the root of the repository, then into `data/`:

```{r read-csv}
ne_counties <- read_csv("./data/ne_counties.csv")
glimpse(ne_counties)
```

This dataset contains demographic and geographic information for counties in Nebraska. Notice that `glimpse()` tells you the number of rows and columns, and shows the data type of each column.


## Part 7: Basic plotting

The simplest way to make a histogram in R is with `hist()`. The `main`, `xlab`, and `ylab` arguments control the title and axis labels:

```{r base-hist}
hist(iris$Sepal.Length,
     main = "Distribution of Sepal Length",
     xlab = "Sepal Length (cm)",
     ylab = "Frequency")
```

The `breaks` argument controls the number of bins:

```{r base-hist-bins}
hist(iris$Sepal.Length,
     breaks = 20,
     main = "Distribution of Sepal Length (more bins)",
     xlab = "Sepal Length (cm)",
     ylab = "Frequency")
```



## Your tasks

Complete each task **completely using R code**. You must show your work. Label your variables sensibly and use comments so I can follow your reasoning. Insert a new code chunk for each task (do not reuse the walkthrough chunks above).

### Task 1: R as a calculator

1.1 Create a variable that stores a number representing your approximate age in years. Convert it to days and print the result. (Note: a fake or rounded age is fine.)

1.2 Create two numeric vectors, each of length 5. Add them together, then compute the `mean()` and `sum()` of the result.

### Task 2: Working with data types

2.1 Create a character vector containing the names of four U.S. states. Create a logical vector of the same length indicating whether each state is in the Midwest. Print both vectors.

2.2 Attempt to compute the `mean()` of your character vector from 2.1. Show the result (including any error or warning message) and write a comment explaining what happened and why.

### Task 3: Packages

3.1 In a code chunk, load the `tidyverse` package. Write a comment in your code explaining the difference between `install.packages()` and `library()`, and why you should not include `install.packages()` inside a script you run regularly.

3.2 Run `sessionInfo()` and leave the output visible in your knitted document.

### Task 4: Reading and exploring a dataset

4.1 Read the `ne_counties.csv` file using `read_csv()`. Run `head()`, `tail()`, and `summary()` on it. For each function, write a short comment describing what it tells you about the data.

4.2 Use `glimpse()` to inspect the dataset. In a comment, report how many rows and columns it has and identify at least one column that is numeric and one that is a character type.

### Task 5: RMarkdown practice

5.1 Insert a **new** code chunk (use the Insert button or keyboard shortcut — do not copy and paste an existing chunk). In that chunk, write code that prints the message `"Lab 00 complete"`. Knit your document and confirm the output appears in the PDF.

### Task 6: Basic plotting

6.1 Using the `ne_counties` data you read in Task 4, make a histogram of the `ALAND10` column (land area) using `hist()`. Give the plot an informative title and label both axes.

6.2 Make a second histogram of a different numeric column from `ne_counties` of your choosing. Use the `breaks` argument to change the number of bins. Label your axes and give the plot a title.


## Questions

Answer the following questions in your `.Rmd` document as prose (plain text, not in a code chunk). There are no wrong answers to question 3.

1. What is the difference between `install.packages()` and `library()`? Why would it be problematic to call `install.packages()` every time you run your script or knit your document?

2. What does a **relative file path** mean? Why does this course require you to use relative paths instead of absolute paths when referencing data files?

3. What did you find confusing or surprising about RStudio, R, or GitHub during this lab? What questions do you still have going into Lab 01?
